---
title: "In_class_activity_Week6"
output: pdf_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# In-class Activity Week 6

```{r}
library(tidyverse)
cherry_data <- read_csv("cherry-bloom-early.csv")
```

## Activity 2

Compute Kendall's Tau between Jan and Feb

```{r}
ken_tau <- cor(cherry_data$avg_tmax_Jan, cherry_data$avg_tmax_Feb, method = "kendall")
print(ken_tau)
```

The estimated Kendall's Tau correlation is 0.317

## Activity 3

Test correlation in 2 by permutation

```{r}
set.seed(123)

# wrap the thing as a function
permute_and_kendall <- function(x, y, reps = 1000$ {
  # catch and throw some errors
  stopifnot(length(x$ == length(y))
  stopifnot(is.numeric(x), is.numeric(y))
  
  # store permutated correlations
  perm_taus <- numeric(reps)
  for $b in seq_len(reps)$ {
    perm_y <- sample(y, replace = FALSE$  # permute y values by setting replace equal false
    perm_taus[b] <- cor(x, perm_y, method = "kendall")
  }
  return(perm_taus)
}

# get perm taus
perm_taus <- permute_and_kendall(cherry_data$avg_tmax_Jan, cherry_data$avg_tmax_Feb)
```

```{r}
hist(perm_taus, breaks=20)
abline(v = ken_tau, lwd = 2)

# compute sampling distribution
print(mean(ken_tau < abs(perm_taus)))
```

We can see the empirical p-value is 0, and it is easy to see that no simulated correlation coefficients in the 1000 experiments actually reaches 0.317. We can hereby reject the null hypothesis of no correlation using the permutation test results and conclude that the correlation of 0.317 between January max temperature and February max temperature is significant.

## Activity 4

```{r}
set.seed(123)

# first create contingency table
con_tab <- table(cherry_data$early_bloom, cherry_data$location)
print(con_tab)

# get test results
pearson_obs <- suppressWarnings(chi$statistic)

reps <- 10000
pearson_perm <- numeric(reps)

for $b in seq_len(reps)$ {
  y <- cherry_data$early_bloom
  # permutate early bloom
  y_perm <- sample(y)
  con_tab_b <- table(y_perm, cherry_data$location)
  
  # pearson's test
  pearson <- chisq.test(con_tab_b, correct = FALSE)
  
  # store perm results
  pearson_perm[b] <- suppressWarnings(pearson$statistic)
}

# permutation empirical p-value $right tail since larger chi-square = more association)
p_perm_pearson <- mean(pearson_perm >= pearson_obs)

p_perm_pearson
```

```{r}
# fisher's exact test
fish <- fisher.test(con_tab)
fish
```

We can see neither Pearson's Chi square test $using permutations$ nor Fisher's exact test $without permutations$ returns significant result. So we cannot reject the null hypothesis that there is no correlation between location and early blossom.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

## Activity 5

```{r}
# helper function to compute quantity of interst theta hat
theta_hat <- function(x$ {
  mu_hat <- mean(x)
  sd_hat <- sd(x)
  return(sd_hat / mu_hat)
}

# helper functio nto generate one sample of X where X = ||Y||_1, Y ~ N(0, I_p)
generate_X_sample <- function(n = 50, p = 5$ {
  Y <- matrix(rnorm(n * p), nrow = n, ncol = p)
  #print(rowSums(abs(Y)))
  return(list(Y = Y, X = rowSums(abs(Y))))
}
```

### 5.1 MC sampling

```{r}
set.seed(123)

M <- 10000
n <- 50
p <- 5

theta_hat_mc <- numeric(M)

for $m in seq_len(M)$ {
  x <- generate_X_sample(n = n, p = p)$X
  theta_hat_mc[m] <- theta_hat(x)
}

mc_mean <- mean(theta_hat_mc)
mc_var  <- var(theta_hat_mc)

# plot MC theta hat
hist(theta_hat_mc, breaks = 20)

print(paste0("mean and variance of theta hat in MC: ", mc_mean, "; ",  mc_var))
```

### 5.2: one bootstrap

```{r}
set.seed(123)

# generate one sample
x0 <- generate_X_sample(n = n, p = p)$X
theta_hat0 <- theta_hat(x0)

# run the bootstrap resampling
B <- 1000
theta_hat_boot <- numeric(B)
for $b in seq_len(B)$ {
  x_boot <- sample(x0, size = n, replace = TRUE$  # bootstrap by setting replace to true
  theta_hat_boot[b] <- theta_hat(x_boot)
}

# plot bootstrap theta hat
hist(theta_hat_boot, breaks = 20)
abline(v = theta_hat0, lwd = 2)

boot_mean <- mean(theta_hat_boot)
boot_var  <- var(theta_hat_boot)

print(paste0("theta_hat for sample: ", theta_hat0))

print(paste0("mean and variance of theta hat in bootstrapping: ", boot_mean, "; ",  boot_var))
```

Observe: looks like the $theta_hat$ in bootstrapping has a slightly smaller mean and slightly smaller variance than in the MC simulations. This could mean the estimation is biased.

### 5.3: repeat bootstrapping

```{r}
set.seed(123)

reps <- 1000
B <- 1000

boot_var_1000 <- numeric(reps)

for $r in seq_len(reps)$ {
  # generate one sample
  x0 <- generate_X_sample(n = n, p = p)$X
  theta_hat0 <- theta_hat(x0)
  
  # run the bootstrap resampling
  B <- 1000
  theta_hat_boot <- numeric(B)
  for $b in seq_len(B)$ {
    x_boot <- sample(x0, size = n, replace = TRUE$  # bootstrap by setting replace to true
    theta_hat_boot[b] <- theta_hat(x_boot)
  }
  
  # store the variance of bootstrapping estimates
  boot_var_1000[r] <- var(theta_hat_boot)
}

# As far as I understand it we should take the mean of these variances?
mean_boot_var <- mean(boot_var_1000)

print(paste0("mean variance of 1000 repeated bootstrapping: ", mean_boot_var))
```

We can see that this time the variance is closer to the MC estimate.

## Activity 6

Here we don't resample from the data, instead we estimate an distribution and draw from that distribution, equivalent to sampling from a empirical distribution

```{r}
# estimate mean and covariance matrix from Y
set.seed(123)

# we already generated one sample, now we estimate its parameters
samples <- generate_X_sample(n = n, p = p)
Y <- samples$Y
X <- samples$X

mu_hat <- colMeans(Y)
Sigma_hat <- cov(Y)
theta_obs <- theta_hat(X)

# bootstrap from the estimated distribution
B <- 1000
theta_hat_para <- numeric(B)

library(MASS)

for $b in seq_len(B)$ {
  # sample from normal
  Y_star <- mvrnorm(n = n, mu = mu_hat, Sigma = Sigma_hat)
  x_star <- rowSums(abs(Y_star))
  theta_hat_para[b] <- theta_hat(x_star)
}

print(paste0("theta from sample: ", theta_obs))
print(paste0("estimated mean: ", mean(theta_hat_para)))
print(paste0("estimated variance: ", var(theta_hat_para)))

hist(theta_hat_para, breaks = 20)
abline(v = theta_obs, lwd = 2)
```

## Activity 7

Theoretical value of $\theta$:

Since $Y \sim N_p(0, I)$ and 

$$
X=|Y|*1=\sum*{j=1}^p |Y_j|.
$$

Each $Y_j \sim N(0,1))$, so $|Y_j|$ is half-normal and have

$$
\mathbb{E}|Z|=\sqrt{\frac{2}{\pi}},\qquad \mathrm{Var}(|Z|)=1-\frac{2}{\pi}.
$$

Then $X$ is a sum of $p$ i.i.d. $|Y|$, so that

$$
\mu_X=\mathbb{E}[X]=p\sqrt{\frac{2}{\pi}},\qquad
\sigma_X^2=\mathrm{Var}(X)=p\Big(1-\frac{2}{\pi}\Big).
$$ 

So the value of interset has a theoretical value:

$$
\theta=\frac{\sigma_X}{\mu_X}
= \frac{\sqrt{p(1-2/\pi)}}{p\sqrt{2/\pi}}
= \frac{\sqrt{\pi/2 - 1}}{\sqrt{p}}.
$$

For p=5: 

$$
\theta = \frac{\sqrt{\pi/2 - 1}}{\sqrt{5}} \approx 0.3378746.
$$


### compute bias from different estimates

```{r}
# 7.1 MC bias estimates
theta_true <- sqrt(pi/2 - 1) / sqrt(p)
bias_mc <- mean(theta_hat_mc) - theta_true

# 7.2 non-parametric bootstrap bias estimates
bias_np  <- mean(theta_hat_boot) - theta_hat0

# 7.3 parametric bootstrap bias estimates
bias_par <- mean(theta_hat_para) - theta_obs  # ideally theta_obs == theta_hat0

#
print(paste0("bias from MC: ", bias_mc, "bias from non-parametric bootstrap: ", bias_np, "bias from parametric bootstrap: ", bias_par))
```

We can see parametric bootstrap has the biggest bias in terms of magnitude, then non-parametric bootstrap, and finally MC has the smallest bias in magnitude.


## Activity 8

### 8.1 Just get 1 CI for each

```{r}
# a function to compute student CI
studentized_ci_from_t <- function(theta_obs, se_obs, t_stats, alpha = 0.05) {
  # get the quantile of bootstrap sampling distribution
  q <- unname(quantile(t_stats, probs = c(alpha/2, 1 - alpha/2)))
  # return inteval
  return(c(theta_obs - se_obs * q[2], theta_obs - se_obs * q[1]))
}

# B2 is the inner loop for calculating 
compute_all_CIs <- function(n = 50, p = 5, B = 1000, B2 = 200, alpha = 0.05) {

  samp <- generate_X_sample(n, p)
  Y <- samp$Y
  x0 <- samp$X
  theta_obs <- theta_hat(x0)
  
  # ----------------------------------------------------------------------------
  # nonparametric bootstrap
  theta_hat_boot <- numeric(B)
  x_boot_list <- vector("list", B)
  
  # percentile CI
  for (b in seq_len(B)) {
    x_b <- sample(x0, n, replace = TRUE)
    x_boot_list[[b]] <- x_b
    theta_hat_boot[b] <- theta_hat(x_b)
  }
  ci_np_percentile <- unname(quantile(theta_hat_boot, probs = c(alpha/2, 1 - alpha/2)))
  
  # studentized CI
  se_obs_np <- sd(replicate(B2, theta_hat(sample(x0, n, replace = TRUE))))
  t_np <- numeric(B)
  for (b in seq_len(B)) {
    x_b <- x_boot_list[[b]]
    # inner loop to compute SE(theta^) (sqrt(V(F^)))
    se_b <- sd(replicate(B2, theta_hat(sample(x_b, n, replace = TRUE))))
    t_np[b] <- (theta_hat_boot[b] - theta_obs) / se_b
  }
  ci_np_student <- studentized_ci_from_t(theta_obs, se_obs_np, t_np, alpha)

  # ----------------------------------------------------------------------------
  # parametric bootstrap
  
  # estimate distribution
  mu_hat <- colMeans(Y)
  Sigma_hat <- cov(Y)

  theta_hat_param <- numeric(B)
  Y_star_list <- vector("list", B)
  
  # percentile CI
  for (b in seq_len(B)) {
    Y_b <- mvrnorm(n, mu_hat, Sigma_hat)
    Y_star_list[[b]] <- Y_b
    theta_hat_param[b] <- theta_hat(rowSums(abs(Y_b)))
  }

  ci_par_percentile <- unname(quantile(theta_hat_param, probs = c(alpha/2, 1 - alpha/2)))
  
  
  # studentized CI
  se_obs_par <- sd(replicate(B2, {
    Y_j <- mvrnorm(n, mu_hat, Sigma_hat)
    theta_hat(rowSums(abs(Y_j)))
  }))

  t_par <- numeric(B)
  for (b in seq_len(B)) {
    Y_b <- Y_star_list[[b]]
    mu_b <- colMeans(Y_b)
    Sigma_b <- cov(Y_b)

    se_b <- sd(replicate(B2, {
      Y_bj <- mvrnorm(n, mu_b, Sigma_b)
      theta_hat(rowSums(abs(Y_bj)))
    }))

    t_par[b] <- (theta_hat_param[b] - theta_obs) / se_b
  }
  ci_par_student <- studentized_ci_from_t(theta_obs, se_obs_par, t_par, alpha)

  list(
    theta_obs = theta_obs,
    ci_np_percentile = ci_np_percentile,
    ci_np_student = ci_np_student,
    ci_par_percentile = ci_par_percentile,
    ci_par_student = ci_par_student
  )
}
```


```{r}
alpha <- 0.05
n <- 50
p <- 5
B <- 1000
B2 <- 200

CIs <- compute_all_CIs(n, p, B, B2, alpha)
CIs

contains <- function(ci, val) ci[1] <= val && val <= ci[2]

c(
  np_pct  = contains(CIs$ci_np_percentile, theta_true),
  np_t    = contains(CIs$ci_np_student, theta_true),
  par_pct = contains(CIs$ci_par_percentile, theta_true),
  par_t   = contains(CIs$ci_par_student, theta_true)
)
```

We can see the 4 CIs all contain the true value

### 8.2 Repeat CIs 1000 times and check coverage

```{r}
set.seed(123)

R <- 1000
cover <- matrix(FALSE, nrow = R, ncol = 4)
colnames(cover) <- c("nonparametric_percentile", "nonparametric_student", "parametric_percentile", "parametric_student")

for (r in seq_len(R)) {
  CIs_1000 <- compute_all_CIs(n, p, B, B2, alpha)
  cover[r, "nonparametric_percentile"]  <- contains(CIs_1000$ci_np_percentile, theta_true)
  cover[r, "nonparametric_student"]    <- contains(CIs_1000$ci_np_student, theta_true)
  cover[r, "parametric_percentile"] <- contains(CIs_1000$ci_par_percentile, theta_true)
  cover[r, "parametric_student"]   <- contains(CIs_1000$ci_par_student, theta_true)
}

cover
```


## Activity 9

```{r}
bca_ci_jacknife_after_bootstrap <- function(x0, B = 1000, alpha = 0.05) {
  n <- length(x0)
  theta_obs <- theta_hat(x0)

  # store bootstrap thetas
  theta_star <- numeric(B)
  counts <- matrix(0L, nrow = B, ncol = n)

  for (b in seq_len(B)) {
    idx <- sample.int(n, size = n, replace = TRUE)
    theta_star[b] <- theta_hat(x0[idx])

    counts[b, ] <- tabulate(idx, nbins = n)
  }

  # z0-----
  z0 <- qnorm(mean(theta_star < theta_obs))

  # jackknife-after-bootstrap leave-one-out
  theta_loo_jab <- numeric(n)
  for (i in seq_len(n)) {
    keep <- counts[, i] == 0L          # replicates where obs i is not included
    theta_loo_jab[i] <- mean(theta_star[keep])
  }
  theta_bar <- mean(theta_loo_jab)

  # acceleration a
  num <- sum((theta_bar - theta_loo_jab)^3)
  den <- 6 * (sum((theta_bar - theta_loo_jab)^2))^(3/2)
  a <- num / den

  # quantile levels
  z_low <- qnorm(alpha/2)
  z_high <- qnorm(1 - alpha/2)

  adj <- function(z) pnorm(z0 + (z0 + z) / (1 - a * (z0 + z)))

  a1 <- adj(z_low)
  a2 <- adj(z_high)

  # bca ci from bootstrap quantiles -----
  ci <- unname(quantile(theta_star, probs = c(a1, a2), type = 7))

  list(ci = ci, z0 = z0, a = a, theta_obs = theta_obs)
}
```

```{r}
set.seed(123)
out <- bca_ci_jacknife_after_bootstrap(x0, B = 1000, alpha = 0.05)
out$ci
```

```{r}
set.seed(123)
R <- 1000
theta_true <- sqrt(pi/2 - 1) / sqrt(5)

cover_bca <- logical(R)

for (r in seq_len(R)) {
  samp <- generate_X_sample(n = 50, p = 5)
  x0 <- samp$X

  bca <- bca_ci_jacknife_after_bootstrap(x0, B = 1000, alpha = 0.05)$ci
  cover_bca[r] <- (bca[1] <= theta_true && theta_true <= bca[2])
}

mean(cover_bca)
```

We can see that the bca coverage 
